
#Final Project Report - Practical Machine Learning Course

##1. Project Goal

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. 

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

##2. Data Sources
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

##3. Analysis and Prediction
First we load the libraries needed for the analysis
```{r results='hide', message=FALSE, warning=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
```
Finally, load the same seed with the following line of code:
```{r}
set.seed(123)
```
###3.1 Getting the data
The training data set can be found on the following URL:

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```

The testing data set can be found on the following URL:

```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

###3.2 Load data to memory 

```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

Partioning the training set for cross validation
Partioning Training data set into two data sets, 60% for crossTraining, 40% for crossTesting:

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
crossTraining <- training[inTrain, ]; crossTesting <- training[-inTrain, ]
dim(crossTraining); dim(crossTesting)
```

###3.3 Cleaning the data
The following transformations were used to clean the data:

Transformation 1: Cleaning NearZeroVariance Variables.

```{r}
nzvData <- nearZeroVar(crossTraining, saveMetrics=TRUE)
```
Use the following code to create another subset without NZV variables:

```{r}
nzvCrossTraining <- names(crossTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")

crossTraining <- crossTraining[!nzvCrossTraining]

dim(crossTraining)
```
Transformation 2: Remove first column of Dataset, which is ID column

```{r}
crossTraining <- crossTraining[c(-1)]
```
Transformation 3: Cleaning Variables with the threshold of 70% of NA's:
```{r}
#Copy crossTraining
training3 <- crossTraining
for(i in 1:length(crossTraining)) { 
        #if num of NAs > 70% of total observations
        if( sum( is.na( crossTraining[, i] ) ) /nrow(crossTraining) >= .7 ) { 
        for(j in 1:length(training3)) {
            #if the columns are the same, remove the column
            if( length( grep(names(crossTraining[i]), names(training3)[j]) ) ==1)  {
                training3 <- training3[ , -j]
            }   
        } 
    }
}
```   
```{r}
dim(training3)
```
```{r}
crossTraining <- training3
rm(training3)
```
Perform transformations on crossTesting and testing data sets.
```{r}
clean1 <- colnames(crossTraining)
clean2 <- colnames(crossTraining[, -58])
crossTesting <- crossTesting[clean1]
testing <- testing[clean2]
```

Examine the result:

```{r}
dim(crossTesting)

dim(testing)
```
Coerce the data into the same type in order to ensure proper functioning of Decision Trees and especially RandomForest Algorithm with the Test data set.
```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(crossTraining)) {
        if( length( grep(names(crossTraining[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(crossTraining[i])
        }      
    }      
}

testing <- rbind(crossTraining[2, -58] , testing)
testing <- testing[-1,]
```
Using Decision Tree for prediction: 
```{r}
modelFitDT <- rpart(classe ~ ., data=crossTraining, method="class")
```
View the decision tree with fancy:
```{r}
fancyRpartPlot(modelFitDT)
```


###3.4 Predicting:
Use the model created from Decision Tree for prediction:
```{r}
predictionsDT <- predict(modelFitDT, crossTesting, type = "class")
```
Use confusion Matrix to test results:
```{r}
confusionMatrix(predictionsDT, crossTesting$classe)
```
Using Random Forests for prediction:
```{r}
modelFitRF <- randomForest(classe ~. , data=crossTraining)
```
Predicting us Random Forest:
```{r}
predictionsRF <- predict(modelFitRF, crossTesting, type = "class")
```
Use confusion Matrix to test results:
```{r}
confusionMatrix(predictionsRF, crossTesting$classe)
```

## Result
As we can see from the above, Random Forests yielded better Results.

The prediction result for the quiz is as follows:

```{r}
predict(modelFitRF, testing, type = "class")
```

